A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-30 latex/main.tex

% PARA: we compared against the UShER tree in a bunch of ways
% NOTE: Trying to clarify that this of the pandemic period and we're 
% not trying to cover *all* history.
The ARG is a comprehensive record of the pandemic phase 
of SARS-CoV-2 evolutionary history, and accurately reflects
a range of results from the literature.
First, we compared the sc2ts ARG and with the state-of-the-art UShER 
tree inferred from the same dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites
(STAR Methods)
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the UShER tree and 
a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We validated that the broad phylogenetic structure of the ARG and 
UShER tree closely match by generating tanglegrams for a set of representative
samples from 1097 deeply-sampled lineages (STAR Methods; 
Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
We also evaluated how well the ARG reflects the structure 
of the Pango nomenclature system~\cite{Rambaut2020} by computing 
Pango assignments for all 2.7 million nodes in the ARG using 
Pangolin~\cite{OToole2021assignment} (STAR Methods).
All but 21,680 of the exported sample alignments (0.87\%) agreed
with the source consensus sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).

@@

The ARG itself is a useful resource to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER.
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of this valuable dataset
for a variety of evolutionary and genetic analyses beyond those presented here.
% We hope that the ARG will spur on new research areas,
% in which recombination is naturally and fully accounted for,
% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
So far, recent developments in ARG methods have focused almost 
exclusively on applications to human sequencing data. In that space,
the availability of ARGs reconstructed from large datasets has already
had tremendous impact on improving our understanding of human genetic
ancestry and past demography \cite{Wohns2022}. We expect
similar leaps to be enabled through tailoring ARG reconstruction methods
to data from other species and pathogens. Specifically, existing
inference methods can be adapted to use the sc2ts ARG for the 
genealogy-based study of selection \cite{hejase2022deep} or the
geographic spread of the virus through time
\cite{Grundler2025geographic}. This is in addition to simpler
analyses, such as calculating mutation spectra or querying the number
or frequency of mutations at a given position, which the ARG makes 
significantly more efficient, and ultimately more easily reproducible.
Since the ARG is a condensed representation of the genetic information
contained in the sequencing data, it can also be used to understand,
for instance, how the variability in sequencing efforts across time 
and location affects our ability to reconstruct the origin of 
important lineages and detect new recombinants.
Since reconstructing the ARG is computationally expensive, 
and checking and improving various aspects of its quality is laborious, 
we hope that making this resource openly available will save
significant time and 
effort for other researchers and enable immediate focus on analysing 
the information contained within this data. In-depth analysis of the
ARG is also incredibly useful for identifying important directions for
further research, both regarding methods (where current models do not
correctly capture features of the data) and data processing 
(where gaps in sequencing or data artifacts result in clear
reconstruction errors).

We contribute several novel methods and concepts which may find applications beyond sc2ts.
First, we devise a simplified LS model that extends the classic concept of parsimony.
In sc2ts, this model allows for accelerated HMM likelihood calculations,
enabling reconstruction of the pandemic-scale ARG presented here.
Existing HMMs used in statistical genetics may be modified in a similar way for significant speedup.
Second, we propose the concept of mutations averted by recombination,
which leads to an interpretable quantity to weigh the support for 
the hypothesis that a given sample is a recombinant against the hypothesis that it is a non-recombinant,
considering the sampled genetic diversity in an ARG and a specific HMM.
Third, we propose the concept of HMM cost to identify poorly explained samples
which are either too divergent or have a high number of mutations inconsistent with their collection dates.
Here we use HMM cost to reject time traveling samples,
which no existing phylogenetic method detects and handles during inference
(e.g., in the Viridian UShER phylogeny, parent and child nodes
do not necessarily follow time order \cite{Hunt2024}),
resulting in a cleaner dataset without samples
with wrong collection dates that may affect downstream analyses.
Future work may be estimating dates
for the suspected time traveling samples using the ARG provided here and
the sc2ts sample matching engine.

@@

We used Pango lineage designations to select representative samples
for the majority of the 2026 Pango lineages in the ARG.
We first identify an ``origination node'' for each Pango lineage,
defined as the oldest node in the ARG with that Pango designation
whose descendants in any tree include more than 50\% of the Pango samples of
the focal type.
The representative sample is then chosen as the oldest sample descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We ``simplify''\cite{Kelleher2018efficient,Wong2024} 
the ARG to include only the representative samples,
resulting in an intermediate ARG of 1867 samples,
then additionally remove internal samples,
known recombinants (i.e.\ the Pango X lineages) and three ``singleton recombinant
samples''
in which a single Pango lineage appears immediately below a recombination node.
This results in a simplified ARG of 1667 samples, each with a different Pango designation.

We compare topology of the simplified ARG against the intersection UShER tree
(which has been dated) simplified down to the same 1667 samples.

The simplified ARG contains four single-breakpoint recombination nodes,
corresponding to the origins of Delta, BA.2, BA.5, and a further BA.2 recombinant
which we deem artifactual by our standard criteria
and treat as non-recombinant for the purposes of plotting
(the breakpoint is on the far right at position 27382,
and supported by only 4 distinguishing sites / 3 loci).
To reduce plot density, we consider only those Pango lineages with fewer than 10 samples,
and extract non-recombinant structures from the ARG
by creating a ``base tree'' of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, and BA.5),
a Delta subtree of 160 Pangos,
a BA.2 subtree of 151 sample (excluding BA.5 descendants) and
a BA.5 subtree of 251 samples.
Figure SX shows each of these trees compared against the equivalent dated UShER topology,
with the left and right parents of Delta and BA.2 highlighted in the base tree.
The order of tips in the two trees was chosen to reduce tangling,
as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012},
with tanglegrams plotted using the SVG outputting facility built into tskit.
The entire pipeline is saved as a notebook at URL (?).

@@

\subsubsection*{Calculation of Pango lineage distance using pangonet}
We define a ``pseudo-phylogenetic'' distance to quantify the divergence between two Pango lineages.
The idea is to count the number of edges separating two given Pango lineages in a ``pseudo-phylogeny''
which relates all the designated Pango lineages.
To construct this pseudo-phylogeny, we used \textit{pangonet},
which mines the information on the Pango lineages in the ``alias\_key.json''
and ``lineage\_notes.txt'' files
from the Pango designation issues (https://github.com/cov-lineages/pango-designation).
If two Pango lineages are siblings in the pseudo-phylogeny,
the distance is defined as the sum of the number of edges from each lineage to their MRCA;
if they are non-siblings (i.e. one Pango is a descendant of the other),
the distance is defined as the number of edges separating them;
if they are identical, the distance is defined as zero.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed a more comprehensive mutational landscape
of SARS-CoV-2 by mapping deletions onto the ARG
using post-hoc parsimony.
We identified major deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All these deletions are known to occur in major VOCs
(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
\cite{Akaishi2023,Li2023indels}).
Eight of these deletions occur in S1-NTD,
a known hotspot for recurrent deletions \cite{McCarthy2021}.

The placement of the following major deletion events
in the ARG matches the phylogenetic placement of
VOC-associated deletion events
suggested in a previous study \cite{Akaishi2023}:
a 3-base deletion 21,991 above the Alpha origin node;
two 6-base deletions at 22,029 and 28,248 above the Delta origin node;
a 9-base deletion at 28,362 above Omicron (B.1.1.529) origin node;
a 3-base deletion at 6,513,
a 9-base deletion at 21,987, and
a 3-base deletion at 22,194 above the BA.1 origin node;
a 9-base deletion at 21,633 above the BA.2 origin node.

However, the placement of several major deletion events
in the ARG differs from the phylogenetic placement of
the corresponding deletion events suggested
in Akaishi and Fujiwara \cite{Akaishi2023}.
That study \cite{Akaishi2023} suggests that
a 6-base deletion at 21,765 occurred
before the divergence of Alpha and Omicron,
but in the ARG this deletion appears to be recurrent,
occurring above the origin nodes of Alpha, BA.1, and BA.4.
Similarly, that study \cite{Akaishi2023} suggests that
a 1-base deletion at 28,271 occurred
before the divergence of Alpha, Delta, and Omicron,
but it is mapped as a recurrent deletion in the ARG
(at the Alpha origin node and above node 1436808).

% This is a separate paragraph discussing where the model has 
% broken down a bit.
Furthermore, there are three contiguous deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node (starting at 11,283) and
the other above the BA.2 origin node (starting at 11,288)
in the ARG (Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up during deletion remapping
because they had four overlapping sites at 11,288-11,292
(which was mapped as the 4-base deletion event).
Furthermore, the analysis of
Akaishi and Fujiwara \cite{Akaishi2023} suggests that
a 9-base deletion at 11,288
occurred before the divergence of Alpha and Omicron.
In the ARG, however, this deletion is mapped
above the Alpha origin node.

By mapping deletion events on the ARG,
we also identified highly recurrent deletions,
for example, the top 20 deletions in terms of
the number of occurrences in the ARG
shown in Table~\ref{tab:rec_dels}.
About half of these deletions are 1-base and
occur in the 5' or 3' UTR of the SARS-CoV-2 genome
(from the start to 250 and from 29,700 to the end),
which are enriched for poor-quality sequencing data.
Genuine recurrent deletions are captured in the ARG,
for example, two known recurrent deletions in S1-NTD
associated with antibody escape \cite{McCarthy2021}.
There are also 1-base recurrent deletions
at 29,541, 29,543, and 29,555
in a non-coding region (between N and ORF10)
which are plausibly homoplasic as they might arise due to
polymerase skipping during genome replication and
this regions is likely free from strong selective pressure.
An in-depth analysis of all the recurrent deletions in the ARG
is beyond the scope of this study,
but we believe that the ARG will serve as a rich resource
to facilitate future evolutionary studies of
deletions in SARS-CoV-2.

@@

\begin{table}
\centering
\caption{
Highly recurrent deletions.
These are the top 20 deletions
in terms of the number of occurrences in the ARG.
Abbreviations:
NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12754 \\
29762 & 1 & non-coding & 11383 \\
29779 & 1 & non-coding & 10951 \\
29769 & 1 & non-coding & 10809 \\
29774 & 1 & non-coding & 10735 \\
29700 & 1 & non-coding & 8562 \\
686 & 9 & ORF1ab / NSP1 & 6436 \\
29614 & 1 & ORF10 & 5039 \\
29555 & 1 & non-coding & 4814 \\
29543 & 1 & non-coding & 4804 \\
29541 & 1 & non-coding & 4802 \\
203 & 1 & non-coding & 3604 \\
222 & 1 & non-coding & 3442 \\
21991 & 3 & S / S1-NTD & 2088 \\
28254 & 1 & ORF8 & 1969 \\
28093 & 1 & ORF8 & 900 \\
21765 & 6 & S / S1-NTD & 482 \\
27532 & 1 & ORF7a & 404 \\
335 & 1 & ORF1ab / NSP1 & 370 \\
337 & 1 & ORF1ab / NSP1 & 346 \\
\bottomrule
\end{tabular}
\end{table}

@@

We performed an indepth analysis of the sc2ts and UShER ARGs 
(we refer to the UShER tree as an ARG for simplicity here as a single 
tree is also an ARG) in terms of parsimony.
A direct comparison is difficult because of the different aligments 
used as input, the different sets of sites used for inference and 
the fact that we are including deletions for a subset of sites in
the sc2ts ARG.

>>>>

% PARA: we compared against the UShER tree in a bunch of ways
% NOTE: Trying to clarify that this of the pandemic period and we're 
% not trying to cover *all* history.
The ARG is a comprehensive record of the pandemic phase 
of SARS-CoV-2 evolutionary history, and accurately reflects
a range of results from the literature.
First, we compared the sc2ts ARG and with the state-of-the-art UShER 
tree inferred from the same dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites
(STAR Methods)
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the UShER tree and 
a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We validated that the broad phylogenetic structure of the ARG and 
UShER tree closely match by generating tanglegrams for a set of representative
samples from 814 deeply-sampled lineages (STAR Methods; 
Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
We also evaluated how well the ARG reflects the structure 
of the Pango nomenclature system~\cite{Rambaut2020} by computing 
Pango assignments for all 2.7 million nodes in the ARG using 
Pangolin~\cite{OToole2021assignment} (STAR Methods).
All but 21,680 of the exported sample alignments (0.87\%) agreed
with the source consensus sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).

@@

The ARG itself is a useful resource to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
which are shown in different studies
using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
The ARG displays remarkable phylogenetic congruence
with a state-of-the-art phylogeny built using UShER.
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of this valuable dataset
for a variety of evolutionary and genetic analyses beyond those presented here.
% We hope that the ARG will spur on new research areas,
% in which recombination is naturally and fully accounted for,
% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
So far, recent developments in ARG methods have focused almost 
exclusively on applications to human sequencing data.
In that space, the availability of ARGs reconstructed from large datasets
has already had tremendous impact on improving our understanding of
human genetic ancestry and past demography \cite{Wohns2022}.
We expect similar leaps to be enabled
through tailoring ARG reconstruction methods
to data from other species and pathogens.
Specifically, existing inference methods can be adapted
to use the sc2ts ARG for the genealogy-based study of selection \cite{hejase2022deep}
or the geographic spread of the virus through time \cite{Grundler2025geographic}.
This is in addition to simpler analyses,
such as calculating mutation spectra
or querying the number or frequency of mutations at a given position,
which the ARG makes significantly more efficient, and
ultimately more easily reproducible.
Since the ARG is a condensed representation of the genetic information
contained in the sequencing data, it can also be used to understand,
for instance, how the variability in sequencing efforts across time 
and location affects our ability to reconstruct the origin of 
important lineages and detect new recombinants.
Since reconstructing the ARG is computationally expensive, 
and checking and improving various aspects of its quality is laborious, 
we hope that making this resource openly available will save significant time
and effort for other researchers and enable immediate focus on analyzing
the information contained within this data. In-depth analysis of the
ARG is also incredibly useful for identifying important directions for
further research, both regarding methods (where current models do not
correctly capture features of the data) and data processing 
(where gaps in sequencing or data artifacts result in clear
reconstruction errors).

We contribute several novel methods and concepts
which may find applications beyond sc2ts.
First, we devise a simplified LS model that extends the classic concept of parsimony.
In sc2ts, this model allows for accelerated HMM likelihood calculations,
enabling reconstruction of the pandemic-scale ARG presented here.
Existing HMMs used in statistical genetics may be modified
in a similar way for significant speedup.
Second, we propose the concept of mutations averted by recombination,
which leads to an interpretable quantity to weigh the support for 
the hypothesis that a given sample is a recombinant against
the hypothesis that it is a non-recombinant,
considering the sampled genetic diversity in an ARG and a specific HMM.
Third, we propose the concept of HMM cost
to identify poorly explained samples
which are either too divergent or
have a high number of mutations inconsistent with their collection dates.
Here we use HMM cost to reject time-traveling samples,
which no existing phylogenetic method detects and handles during inference
(e.g., in the Viridian UShER phylogeny, parent and child nodes
do not necessarily follow time order \cite{Hunt2024}),
resulting in a cleaner dataset without samples
with wrong collection dates that may affect downstream analyses.
Future work may be to estimate dates
for the suspected time traveling samples using the ARG provided here and
the sc2ts sample matching engine.

@@

We compare the non-recombinant parts of the ARG with the UShER topology
by simplifying the phylogenies to a smaller number of
Pango representative samples.
We focus on the 1473 Pango lineages that have perfect ``origination events'' 
as described in \ref{sec:supplement_pango_lineages}, and
omit Pango designations starting with "X",
to avoid recombinants, leaving 1329 Pango lineages.
To ensure that each sample is a tip (rather than an internal node)
in the simplified topologies,
a representative sample is then chosen as the oldest sample
descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We then further remove 4 samples
which are singleton descendants of a recombination event in the ARG
(representing potentially missed Pango recombinant lineages),
as well as 43 samples that are descendants of the recombinant XBB
but which do not start with an "X".
This results in a simplified ARG of 1282 samples,
each sample having a different Pango designation.

The simplified ARG contains two single-breakpoint recombination nodes. The first
corresponds to a BA.2 recombinant whose breakpoint is on the far right (position 27382) supported by only 4 sites / 3 loci, anch which therefore does not pass our recombinant quality control measure. The second is the origin of BA.5, which (the ).
To reduce plot density, we consider only those Pango lineages with fewer than 10 samples,
and extract non-recombinant structures from the ARG
by creating a ``base tree'' of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, and BA.5),
a Delta subtree of 160 Pangos,
a BA.2 subtree of 151 sample (excluding BA.5 descendants) and
a BA.5 subtree of 251 samples.
Figure SX shows each of these trees compared against the equivalent dated UShER topology,
with the left and right parents of Delta and BA.2 highlighted in the base tree.
The order of tips in the two trees was chosen to reduce tangling,
as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012},
with tanglegrams plotted using the SVG outputting facility built into tskit.
The entire pipeline is saved as a notebook at URL (?).

@@

\subsubsection*{Calculation of Pango lineage distance using pangonet}
We define a ``pseudo-phylogenetic'' distance
to quantify the divergence between two Pango lineages.
The idea is to count the number of edges
separating two given Pango lineages in a pseudo-phylogeny
which relates all the designated Pango lineages.
To construct this pseudo-phylogeny, we used \textit{pangonet},
which mines the information on the Pango lineages
in the ``alias\_key.json'' and ``lineage\_notes.txt'' files
from the Pango designation issues
(\url{https://github.com/cov-lineages/pango-designation}).
If two Pango lineages are siblings in the pseudo-phylogeny,
the distance is defined as the sum of the number of edges from each lineage to their MRCA;
if they are non-siblings (i.e. one Pango is a descendant of the other),
the distance is defined as the number of edges separating them;
if they are identical, the distance is defined as zero.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed
a more comprehensive mutational landscape of SARS-CoV-2
by mapping deletions onto the ARG using post-hoc parsimony.
We identified major deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All these deletions are known to occur in major VOCs 
(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
\url{https://github.com/cov-lineages/pango-designation/issues/361};
\cite{Stern2021,Akaishi2023,Li2023indels},
except the 4-base deletion and two 5-base deletions in ORF1ab,
which are likely artefacts of deletion remapping (see below).
Notably, eight of these deletions occur
in the N-terminal domain of the S1 unit in Spike,
a known hotspot for recurrent deletions \cite{McCarthy2021}
(also see Table~\ref{tab:rec_dels}).

The placement of the following major deletion events
in the ARG aligns well with the phylogenetic placement of
VOC-associated deletion events
suggested in a previous study \cite{Akaishi2023}:
a 3-base deletion 21,991 above the Alpha origin node;
two 6-base deletions at 22,029 and 28,248 above the Delta origin node;
a 9-base deletion at 28,362 above the B.1.1.529 origin node;
a 3-base deletion at 6,513,
a 9-base deletion at 21,987, and
a 3-base deletion at 22,194 above the BA.1 origin node;
a 9-base deletion at 21,633 above the BA.2 origin node.

However, the ARG suggests that two major deletions
(a 6-base deletion at 21,765 and a 1-base deletion at 28,271;
Table~\ref{tab:major_dels}) are recurrent
(i.e., acquired multiple times),
but Akaishi and Fujiwara \cite{Akaishi2023} suggest that
these deletions might have been acquired
before the divergence of certain VOCs,
based on an analysis of a small set of 20 samples.
That study suggests that the 6-base deletion
occurred before the divergence of Alpha and Omicron,
but in the ARG this deletion occurs three times,
above the origin nodes of Alpha, BA.1, and BA.4.
Similarly, that study suggests that the 1-base deletion
occurred before the divergence of Alpha, Delta, and Omicron,
but in the ARG this deletion occurs
above the Alpha origin node and above node 1436808.

There are three contiguous deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node (starting at 11,283) and
the other above the BA.2 origin node (starting at 11,288)
in the ARG (Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up during deletion remapping
as they had four overlapping sites at 11,288-11,292
(which was mapped as the 4-base deletion event).
Furthermore, Akaishi and Fujiwara \cite{Akaishi2023}
suggest that a 9-base deletion at 11,288 occurred
before the divergence of Alpha and Omicron.
In the ARG, however, this deletion occurs
above the Alpha origin node.

We identified highly recurrent deletions,
for example, the top 20 deletions in terms of
the number of occurrences in the ARG shown in Table~\ref{tab:rec_dels}.
These include known recurrent deletions, for example,
6-base deletion at 21,765,
3-base deletion at 21,991, and
3-base deletion 22,194 in S1-NTD
associated with antibody escape \cite{McCarthy2021,Meng2021}.
The out-of-frame deletions in S1-NTD
(e.g., the 2-base deletion at 21,765),
N, ORF1ab, and ORF8
are implausible and are likely artefacts caused by
sequencing errors or sequence misalignments.
Future improvements in
data pre-processing and deletion remapping
should remove such deletions.

@@

\begin{table}
\centering
\caption{
Highly recurrent deletions.
These are the top 20 deletions
in terms of the number of occurrences in the ARG.
Abbreviations:
NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
686 & 9 & ORF1ab / NSP1 & 6436 \\
21991 & 3 & S / S1-NTD & 2088 \\
21765 & 6 & S / S1-NTD & 482 \\
22029 & 6 & S / S1-NTD & 275 \\
22194 & 3 & S / S1-NTD & 200 \\
6513 & 3 & ORF1ab / NSP3 & 181 \\
21765 & 2 & S / S1-NTD & 140 \\
21995 & 1 & S / S1-NTD & 90 \\
21988 & 3 & S / S1-NTD & 78 \\
21987 & 7 & S / S1-NTD & 78 \\
21993 & 3 & S / S1-NTD & 77 \\
21987 & 9 & S / S1-NTD & 70 \\
21994 & 2 & S / S1-NTD & 65 \\
28370 & 1 & N & 61 \\
11292 & 3 & ORF1ab / NSP6 & 48 \\
21990 & 1 & S / S1-NTD & 46 \\
21987 & 4 & S / S1-NTD & 46 \\
21990 & 6 & S / S1-NTD & 45 \\
686 & 7 & ORF1ab / NSP1 & 43 \\
28252 & 2 & ORF8 & 41 \\
\bottomrule
\end{tabular}
\end{table}

@@

We performed an in-depth analysis of the sc2ts and UShER ARGs
(we refer to the UShER tree as an ARG for simplicity here,
as a single tree is also an ARG) in terms of parsimony.
A direct comparison is difficult because of
the different alignments used as input,
the different sets of sites used for inference and
the fact that we are including deletions
for a subset of sites in the sc2ts ARG.
